Longitudinal Principal Manifold Estimation

### 2023-10-11 paper/main.tex
add abstract

@@

The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.

@@

\subsection*{Data Reduction}
We begin by considering $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} (Section 5.3 therein) to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$. HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.

This step reduces the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to the size $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

\subsection*{Initialization}
The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another) \meng{I would suggest writing more explanations for the ``inconsistent parameterization directions." Some visual illustrations are necessary.}. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed. \meng{This paragraph should be devoted to the explanation for ``inconsistent parameterization directions." The comparison can be presented using visualizations.}

Instead of using Isomap on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one Isomap run to reach a parameterization for the centers across all time points simultaneously. Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time. \meng{I would suggest presenting simulation studies showing the ``consistency over time." Specifically, we may compare the difference between time-wise and over-time initializations via simulations.}

\subsection*{Fitting}
Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$, which deals with the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in Eq.~\eqref{eq:11}.

The remainder of the fitting stage deals with the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$, i.e., the curvature in the time dimension (see Eq.~\eqref{eq:11}). Specifically, the remainder of this stage involves estimating a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$. \meng{The term ``coefficient space" is clear to me, but it might be confusing to other readers who are unfair with the details of PME.}

At each time point, the estimated principal manifold has a spline form. As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points being used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ being specified based on the maximum number of clusters identified in the data reduction step at any time point. This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
\begin{itemize}
    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
\end{itemize}
This minimization is conducted by solving
\begin{equation}
  \left(
    \begin{array}{cc}
      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
      \boldsymbol{R} & \boldsymbol{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      \boldsymbol{s}_t \\
      \boldsymbol{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \boldsymbol{Y}_t \\
      \boldsymbol{0}
    \end{array}
  \right). \label{eq:13}
\end{equation}
The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
\begin{equation}
  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
\end{equation}

>>>>

We propose a longitudinal extension to principal manifold estimation, with the goal of recovering smooth, longitudinally meaningful manifold estimates over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We then apply the proposed method to estimating the surfaces of hippocampi of healthy individuals in the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

Given the framework proposed in the previous Section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}). Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping. Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space. For example, the PME algorithm can be used to obtain this parameterization. In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, thus smoothing over the spatial spline coefficients with respect to time. This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.

@@

\subsection*{Sample Size Reduction}

Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see\cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

\subsection*{Initialization}

The next step in developing the algorithm for estimation of the longitudinal principal manifold is initialization. Different algorithms use different approaches for this step. For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, implementation of commonly used approaches for initialization in our setting can be problematic. For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e. the brain region of interest at each time point) is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.

To obtain the same parameterization of the manifold over time, instead of using Isomap on the data for each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously. This approach is based on the assumption that the shape of the structure, i.e. the brain region of interest, remains relatively stable over time. In Section \ref{s:simulations}, we consider some deviations from this assumption and their effect on the results and we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.

\subsection*{Fitting}

Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa}. At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$. The principal manifold $\widehat{f_t}$ at time $t$ is characterized by spline coefficients. Hence, we estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.

At each time point, the estimated principal manifold has a spline form. As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point. This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
\begin{itemize}
    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
\end{itemize}
This minimization is conducted by solving
\begin{equation}
  \left(
    \begin{array}{cc}
      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
      \boldsymbol{R} & \boldsymbol{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      \boldsymbol{s}_t \\
      \boldsymbol{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \boldsymbol{Y}_t \\
      \boldsymbol{0}
    \end{array}
  \right). \label{eq:13}
\end{equation}
The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
\begin{equation}
  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
\end{equation}
